% TODO: goal synonymous with conclusion??!

\def\prod{\mathrel{::=}}
\def\altn{\mathrel{\mid}}
\def\NT#1{\textsf{#1}}
\def\TT#1{\texttt{#1}}
\def\Nat{\mathbb{N}}

\def\Expression{\NT{Expression}}
\def\Add{\oplus}
\def\PUSH{\TT{PUSH}\;}
\def\ADD{\TT{ADD}}
\def\nil{\TT{[]}}
\def\Instruction{\NT{Instruction}}
\def\Code{\NT{Code}}
\def\Stack{\NT{Stack}}
\def\Machine{\NT{Machine}}
\def\compile{\textit{compile}}

\def\Eval{\Downarrow}
\def\Step{\mapsto}
\def\StepS{\Step^\star}
\def\Exec{\rightarrowtail}
\def\ExecS{\Exec^\star}

\chapter{Semantics for Compiler Correctness}\label{ch:semantics}

%\begin{itemize}
%\item denotational, small-step for $(+,N)$
%\item equivalence proof
%\item rule induction (coinduction?)
%\item machine semantics for $\{PUSH m, ADD\} x N$
%\item statement of compiler correctness (style? direct vs CPS)
%\item compiler correctness as running example: proof
%\end{itemize}

In the context of computer science, semantics is the study of the meaning of
programming languages. Having a mathematically rigorous definition of the
language allows us to reason about (programs written in) the language
precisely and without ambiguity. In this chapter, we will take an elementary
look at different ways of giving meaning to a language, and various
techniques for proving properties about programs. After this prelude, we
reach the nub of this chapter, where we shall define a simple compiler from
an expression language to a stack machine, and explore what it means to say
that the compiler is correct.

\section{Semantics}%{{{%

\subsection{Natural Numbers and Addition}%{{{%

To unambiguously reason about what any given program means, we need to give
a mathematically rigorous definition of the language in which it is
expressed. To this end, let us consider the elementary language of natural
numbers and
addition~\cite{hutton04-exceptions,hutton06-calculating,hutton07-interruptions}.
\begin{align*}
	\Expression
		& \prod \Nat \eqTag{Exp-$\mathbb{N}$} \\
		& \altn \Expression \Add \Expression \eqTag{Exp-$\Add$}
\end{align*}
That is, an $\Expression$ is either simply a natural number, or a pair of
$\Expression$s, punctuated with the $\Add$ symbol to represent the
operation of addition. We will adhere to a naming convention of $m, n \in
\Nat$ and $a, b, e \in \Expression$.

Although seemingly simplistic, this language has sufficient structure to
illustrate the essential aspects of computation, namely that of sequencing
computations and combining their results, as we shall expand upon later in
sections \ref{sec:small-step} and \ref{sec:semantics-degenerate}.

%}}}%

\subsection{Denotational Semantics}%{{{%

% semantic brackets
\def\sb[#1]{[\![#1]\!]}

Denotation semantics attempts to give an interpretation of the source
language in some suitable existing formalism that we already understand.
More specifically, the denotation of a program is a representation of what
the program means in the vocabulary of the chosen formalism, which could be
the language of sets and functions, the $\lambda$-calculus, or perhaps one
of the many process calculi. Thus, to formally give a denotational semantics
for a language is to define a mapping from the source language into some
underlying semantic domain.

For example, we can give the following semantics for our earlier
$\Expression$ language, denoted as a natural number:
\begin{align*}
	\sb[\anonymous] &: \Expression \rightarrow \Nat \\
	\sb[ m ] &= m \eqTag{denote-val} \\
	\sb[ a \Add b ] &= \sb[ a ] + \sb[ b ] \eqTag{denote-plus}
\end{align*}
Here, a numeric $\Expression$ is interpreted as just the number itself. The
denotation of the $\Add$ operator is the familiar $+$ on natural numbers;
alternatively, we could say that $a \Add b$ is denoted by the sum of the
denotations of its sub-expressions $a$ and $b$. The expression $\sb[(1
\Add 2) \Add (4 \Add 8)]$ say, has the denotation $15$ by repeatedly
applying the above definition:
\begin{align*}
	\sb[(1 \Add 2) \Add (4 \Add 8)]
		&= \sb[1 \Add 2] + \sb[4 \Add 8] \\
		&= (\sb[1] + \sb[2]) + (\sb[4] + \sb[8]) \\
		&= (1 + 2) + (4 + 8) = 15
\end{align*}

%}}}%

\subsection{Big-Step Operational Semantics}%{{{%

The notion of big-step operational semantics is concerned with the overall
result of a computation. Formally, we define a relation ${\Eval}
: \Expression \times \Nat$ between $\Expression$s and their final values,
given below in a natural deduction style:
\begin{gather*}
\inferrule*{ }
	{m \Eval m} \eqTag{big-val} \\[2ex]
\inferrule*{a \Eval m \quad b \Eval n}%
	{a \Add b \Eval m + n} \eqTag{big-plus}
\end{gather*}
The first \eqName{big-val} rule says that a simple numeric $\Expression$
evaluates to the number itself. The second \eqName{big-plus} rule states
that, if $a$ evaluates to $m$ and $b$ evaluates to $n$, then $a \Add b$
evaluates to the sum $m + n$. Thus according to this semantics, $(1 \Add
2) \Add (4 \Add 8) \Eval 15$ by the following derivation:
\begin{gather*}
\inferrule* [left=big-plus]
{
	\inferrule* [Left=big-plus]
	{
		\inferrule* [Left=big-val]
			{ }{1 \Eval 1}
		\and
		\inferrule*
			{ }{2 \Eval 2}
	}
	{1 \Add 2 \Eval 3}
	\and
	\inferrule*
	{
		\inferrule*
			{ }{4 \Eval 4}
		\and
		\inferrule*
			{ }{8 \Eval 8}
	}
	{4 \Add 8 \Eval 12}
}
{(1 \Add 2) \Add (4 \Add 8) \Eval 15}
\end{gather*}

\noindent One advantage of a relational operational semantics is that the
behaviour can be non-deterministic, in the sense that each expression could
potentially evaluate to multiple distinct values. In contrast,
a denotational semantics deals with non-determinism in the source language
by mapping it to a potentially different notion of non-determinism in the
underlying formalism. Should we require our expression language to be
non-deterministic, we would need to switch the denotational domain of the
previous semantics to the power set of natural numbers, rather than just the
set of natural numbers.

%}}}%

\subsection{Small-Step Operational Semantics}\label{sec:small-step}%{{{%

Small-step semantics on the other hand is concerned with how a computation
proceeds as a sequence of steps. Both big-step and small-step semantics are
`operational' in the sense that the meaning of a program is understood
through how it operates to arrive at the result. However, in this case each
reduction step is made explicit, which is particularly apt when we wish to
consider computations that produce side-effects. Again we formally define
a relation ${\Step} : \Expression \times \Expression$, but between pairs
of $\Expression$s in this instance:
\begin{gather*}
\inferrule*{ }%
	{m \Add n \Step m + n} \eqTag{small-plus} \\[2ex]
\inferrule*{b \Step b'}%
	{m \Add b \Step m \Add b'} \eqTag{small-right} \\[2ex]
\inferrule*{a \Step a'}%
	{a \Add b \Step a' \Add b} \eqTag{small-left}
\end{gather*}
The first rule \eqName{small-plus} deals with the case where the expressions
on both sides of $\Add$ are numerals: in a single step, it reduces to the
sum $m + n$. The second \eqName{small-right} rule applies when the left
argument of $\Add$ is a numeral, in which case the right argument makes
a single reduction, while \eqName{small-left} reduces the left argument of
$\Add$ whenever possible. There is no rule corresponding to a lone numeric
$\Expression$ as no further reductions are possible in this case.

As each ${\Step}$ corresponds to only a single computational step, it will
often be more convenient to refer to it via its reflexive, transitive
closure, defined as follows:
\begin{gather*}
\inferrule*[right=small-nil]
{ }{a \StepS a}
\qquad\qquad
\inferrule*[right=small-cons]
{a \Step a' \and a' \StepS b}
{a \StepS b}
\end{gather*}
Thus, the full reduction sequence of $(1 \Add 2)
\Add (4 \Add 8) \StepS 15$ would begin at the $1 \Add 2$
sub-expression,
\begin{gather*}
\inferrule* [left=small-left]
{
	\inferrule* [Left=small-plus]
	{ }{1 \Add 2 \Step 3}
}
{(1 \Add 2) \Add (4 \Add 8) \Step 3 \Add (4 \Add 8)}
\end{gather*}
followed by $4 \Add 8$,
\begin{gather*}
\inferrule* [left=small-right]
{
	\inferrule* [Left=small-plus]
	{ }{4 \Add 8 \Step 12}
}
{3 \Add (4 \Add 8) \Step 3 \Add 12}
\end{gather*}
before delivering the final sum:
\begin{gather*}
\inferrule* [left=small-plus]
{ }{3 \Add 12 \Step 15}
\end{gather*}

\noindent It would be perfectly reasonable to give a right-to-left, or even
a non-deterministic interleaved reduction strategy for our $\Expression$
language. However, we enforce a left-to-right order to coincide with the
definition of $\bind$ for the $\mathsf{State}\;\alpha$ monad of \S\ref{?},
which we motivate in the following section.

%}}}%

\subsection{Monoids as Degenerate Monads}\label{sec:semantics-degenerate}%{{{%

In the previous chapter, we briefly explored the use of monads as
a mechanism for modelling sequential computations. In the degenerate case
where the result type of the computations form a monoid with an operation
$\cdot$, such computations themselves naturally form a monoid too.
Concretely, suppose we are working in some monad $\mathsf{M}$ computing
values of type $\Nat$, where $(\Nat,\;+,\;0)$ is a monoid. With the
following definition of $\ast$,
\[
	a \ast b \quad=\quad
		a \bind \lambda m \rightarrow
		b \bind \lambda n \rightarrow
		\textit{return}\;(m + n)
\]
such computations themselves form the monoid
$(\mathsf{M}\;\Nat,\;\ast,\;\textit{return}\;0)$. As we have done in
\S\ref{sec:stm-monad-prop}, we can again use equational reasoning to confirm
that the relevant left-/right-identity and associativity properties hold.
Therefore, it is not entirely unreasonable to view monoids as a degenerate
form of monads.

As we only consider values of natural numbers, rather than dealing with
computations of type $\mathsf{M}\;\Nat$, we may as well deal directly with
the underlying $(\Nat,\;+)$ monoid, mirrored in the syntax of the
$\Expression$ language. This simplification allows us to avoid the
orthogonal issues of binding and substitution. By enforcing a left-to-right
evaluation order for $\Add$ to mirror that of $\bind$, we maintain the key
monadic aspect of sequencing computations.

%}}}%

%\TODO{Are the proofs necessary? They're rather elementary.}
%%{{{%
%\begin{verbatim}
%  a * return 0
%={ defn of * }
%  a >>= \ m -> return 0 >>= \ n -> return (m . n)
%={ left-identity of >>= }
%  a >>= \ m -> (\ n -> return (m . n)) 0
%={ application }
%  a >>= \ m -> return (m . 0)
%={ right-identity of . }
%  a >>= \ m -> return m
%={ right-identity of >>= }
%  a
%\end{verbatim}
%%}}}%
%%{{{%
%\begin{verbatim}
%  return 0 * b
%={ defn of * }
%  return 0 >>= \ m -> b >>= \ n -> return (m . n)
%={ left-identity of >>= }
%  (\ m -> b >>= \ n -> return (m . n)) 0
%={ application }
%  b >>= \ n -> return (0 . n)
%={ left-identity of . }
%  b >>= \ n -> return n
%={ right-identity of >>= }
%  b
%\end{verbatim}
%%}}}%
%%{{{%
%\begin{verbatim}
%  (a * b) * c
%={ defn of * }
%  (a >>= \ l -> b >>= \ m -> return (l . m)) >>= \ lm ->
%  c >>= \ n ->
%  return (lm . n)
%={ associativity of >>=, twice }
%  a >>= \ l -> b >>= \ m -> (return (l . m) >>= \ lm ->
%  c >>= \ n ->
%  return (lm . n))
%={ substitute lm by left-id of >>= }
%  a >>= \ l ->
%  b >>= \ m ->
%  c >>= \ n ->
%  return ((l . m) . n)
%={ associativity of . }
%  a >>= \ l ->
%  b >>= \ m ->
%  c >>= \ n ->
%  return (l . (m . n))
%={ factor out (m . n)  }
%  a >>= \ l ->
%  b >>= \ m -> c >>= \ n -> (return (m . n) >>= \ mn ->
%  return (l . mn))
%={ associativity of >>=, twice }
%  a >>= \ l ->
%  (b >>= \ m -> c >>= \ n -> return (m . n)) >>= \ mn ->
%  return (l . mn)
%={ defn of * }
%  a * (b * c)
%\end{verbatim}
%%}}}%

%}}}%

\section{Equivalence Proofs and Techniques}%{{{%

We can now proceed to show various properties of the $\Expression$ language
in a rigorous manner, now that we have provided precise definitions for the
semantics of the language. One obvious questions arises, on the matter of
whether the semantics we have given in the previous section---denotational,
big-step and small-step---agree in some manner. This section reviews the
main techniques for proving such properties.

\subsection{Syntax and Rule Induction}%{{{%

\def\subExp{\sqsubset}

The main tool at our disposal is that of well-founded induction, which we
can apply to any well-founded structure. For example, we can show that the
syntax of the $\Expression$ language satisfies the condition of
well-foundedness when paired with the following sub-expression ordering:
\[
	a \subExp a \Add b \qquad b \subExp a \Add b \eqTag{Exp-$\subExp$}
\]
The partial order given by the transitive closure of $\subExp$ is
well-founded, since any $\subExp$-descending chain of expressions must
eventually end in a numeral at the leaves of the expression tree. This
particular ordering arises naturally from the inductive definition of
$\Expression$: the inductive case \eqName{Exp-$\Add$} allows us to build
a larger expression $a \Add b$ given two existing expressions $a$ and $b$,
while the base case \eqName{Exp-$\mathbb{N}$} constructs primitive
expressions out of any natural number. In this particular case, to give
a proof that some property $P(e)$ holds for all $e \in \Expression$, it
suffices by the well-founded induction principle to show instead that:
\[
	\forall b \in \Expression.\;
		(\forall a \in \Expression.\; a \subExp b \rightarrow P(a))
		\rightarrow P(b)
\]
More explicitly, we are provided with the hypothesis that $P(a)$ already
holds for all sub-expressions $a \subExp b$ when proving $P(b)$; in those
cases when $b$ has no sub-expressions, we must show that $P(b)$ holds
directly.

The application of well-founded induction to the structure of an inductive
definition is called \emph{structural induction}: to prove that a property
$P(x)$ holds for all members $x$ of an inductively defined structure $X$, it
suffices to initially show that $P(x)$ holds in all the base cases in the
definition of $X$, and that $P(x)$ holds in the inductive cases assuming
that $P(x')$ holds for any immediate substructure $x'$ of $x$.

Our earlier reduction rules ${\Eval}$ along with ${\Step}$ and its
transitive closure ${\StepS}$ are similarly inductively defined, and
therefore admits the same notion of structural induction. These instances
will be referred to as \emph{rule induction}.

%}}}%

\subsection{Proofs of Semantic Equivalence}\label{sec:semantic-equivalence}%{{{%

We shall illustrate the above technique with some examples.

\begin{theorem}\label{thm:denote-big}
Denotational semantics and big-step operational semantics coincide:
\[
	\forall e \in \Expression,\ m \in \Nat.\quad
		\sb[e] \equiv m \leftrightarrow e \Eval m
\]
\end{theorem}

\begin{proof}
We consider each direction of the $\leftrightarrow$ biconditional
separately. To show $\sb[e] \equiv m \rightarrow e \Eval m$, we could
proceed by induction on the structure of the definition of the
$\sb[\anonymous]$ function, which happens to be structurally recursive on
its argument. Therefore we may equivalently proceed by structural induction
on $e$, giving us two cases to consider:
\begin{description}
\item[Case $e \equiv n$:]%{{{%
Substituting $e$, this base case specialises to showing that:
\[
	\sb[n] \equiv m \rightarrow n \Eval m
\]
By \eqName{denote-val} in the definition of $\sb[\anonymous]$, the
hypothesis evaluates to $n \equiv m$. This allows us to substitute $m$ for
$n$ in the conclusion, which is trivially satisfied by instantiating
\eqName{big-val} with $m$ in the definition of ${\Eval}$.
%}}}%
\item[Case $e \equiv a \Add b$:]%{{{%
Substituting $e$ as before, we need to show that:
\[
	\sb[a \Add b] \equiv m \rightarrow a \Add b \Eval m
\]
Applying \eqName{denote-plus} once to the hypothesis, we obtain that $\sb[a]
+ \sb[b] \equiv m$. Substituting for $m$, the conclusion becomes $a \Add
b \Eval \sb[a] + \sb[b]$. Instantiate the induction hypothesis twice
with the trivial equalities $\sb[a] \equiv \sb[a]$ and $\sb[b] \equiv
\sb[b]$ to yield proofs of $a \Eval \sb[a]$ and $b \Eval \sb[b]$,
which are precisely the two antecedents required by \eqName{big-plus} to
obtain $a \Add b \Eval \sb[a] + \sb [b]$.
%}}}%
\end{description}

\noindent The second half of the proof requires us to show that $\sb[e]
\equiv m \leftarrow e \Eval m$. We may proceed by structural induction
directly on our assumed hypothesis of $e \Eval m$, which must match
either \eqName{big-val} or \eqName{big-plus} in the definition of
$\Eval$:
\begin{description}
\item[Rule \eqName{big-val}:]%{{{%
Matching $e \Eval m$ with the consequent of \eqName{big-val}, we may
conclude that there exists an $n$ such that $e \equiv n$ and $m \equiv n$.
Substituting $n$ for $e$ and $m$ in $\sb[e] \equiv m$ and applying
\eqName{denote-val} once, the desired conclusion becomes $n \equiv n$, which
is trivially satisfied by the reflexivity of $\equiv$.
%}}}%
\item[Rule \eqName{big-plus}:]%{{{%
Again by matching $e \Eval m$ with the consequent of \eqName{big-plus},
there exists $a$, $b$, $n_a$ and $n_b$ where $e \equiv a \Add b$ and $m
\equiv n_a + n_b$, such that $a \Eval n_a$ and $b \Eval n_b$.
Substituting for $e$ and $m$, the conclusion becomes $\sb[a \Add b] \equiv
n_a + n_b$, which reduces to:
\[
	\sb[a] + \sb[b] \equiv n_a + n_b
\]
by applying \eqName{denote-plus} once. Instantiating the induction
hypothesis twice with $a \Eval n_a$ and $b \Eval n_b$ yields the
equalities $\sb[a] \equiv n_a$ and $\sb[b] \equiv n_b$ respectively, which
allows us to rewrite the conclusion as $\sb[a] + \sb[b] \equiv \sb[a]
+ \sb[b]$ by substituting $n_a$ and $n_b$. The desired result is now
trivially true by reflexivity of $\equiv$.
%}}}%
\end{description}

\noindent Thus we have shown both directions of the theorem.
\end{proof}

% premise, antecedent / consequent
% hypothesis / conclusion

\begin{theorem}\label{thm:big-small}
Big-step and small-step operational semantics coincide. That is,
\[
	\forall e \in \Expression,\ m \in \Nat.\quad
		e \Eval m \leftrightarrow e \StepS m
\]
\end{theorem}
\begin{proof}
We shall consider each direction separately as before. To show the forward
implication, we proceed by rule induction on the assumed $e \Eval m$
hypothesis:
\begin{description}
\item[Rule \eqName{big-val}:]%{{{%
There exists an $n$ such that $e \equiv n$ and $m \equiv n$, by matching $e
\Eval m$ with the consequent of \eqName{big-val}. Substituting $n$ for
both $e$ and $m$, we can readily conclude that $n \StepS n$ via
\eqName{small-nil}.
%}}}%
\item[Rule \eqName{big-plus}:]%{{{%
There exists $a$, $b$, $n_a$ and $n_b$ where $e \equiv a \Add b$ and $m
= n_a + n_b$, such that $a \Eval n_a$ and $b \Eval n_b$. After
substituting for $e$ and $m$, the desired conclusion becomes:
\[
	a \Add b \StepS n_a + n_b
\]
Instantiating the induction hypothesis with $a \Eval n_a$ and $b
\Eval n_b$ gives us evidence of $a \StepS n_a$ and $b
\StepS n_b$ respectively. With the former, we can apply $\anonymous
\Add b$ to each of the terms and \eqName{small-left} to obtain a proof of
$a \Add b \StepS n_a \Add b$, while with the latter, we obtain
$n_a \Add b \StepS n_a \Add n_b$ by applying $n_a \Add
\anonymous$ and \eqName{small-right}.

By the transitivity of $\StepS$, these two small-step reduction
sequences combine to give $a \Add b \StepS n_a \Add n_b$, to
which we need only append an instance of \eqName{small-plus} to arrive at
the conclusion.
%}}}%
\end{description}

\noindent The converse implication $e \Eval m \leftarrow e \StepS m$
additionally requires lemma \ref{lem:small-sound}, which states that $e
\Step e' \rightarrow e' \Eval m \rightarrow e \Eval m$; in other words, the
reduct of a single step under the small-step semantics evaluates under the
big-step semantics to the same value as the original expression. We proceed
by induction over the definition of $\StepS$, given $e \StepS m \rightarrow
e \Eval m$:
\begin{description}
\item[Rule \eqName{small-nil}:]%{{{%
If the reduction sequence is empty, then it follows that $e \equiv m$. In
this case, we can trivially satisfy the conclusion of $m \Eval m$ with
\eqName{big-val}.
%}}}%
\item[Rule \eqName{small-cons}:]%{{{%
For non-empty reduction sequences, there exists an $e'$ such that $e \Step
e'$ and $e' \StepS m$. Invoke lemma \ref{lem:small-sound} with $e
\Step e'$ and $e' \Eval m$, where the latter given by the induction
hypothesis, to obtain proof of $e \Eval m$.
%}}}%
\end{description}
Pending the proof of lemma \ref{lem:small-sound} below, we have thus shown
the equivalence of big- and small-step semantics for the $\Expression$
language.
\end{proof}

\begin{lemma}
\label{lem:small-sound}
A single small-step reduction preserves the value of expressions with
respect to the big-step semantics:
\[
	\forall e, e' \in \Expression,\ m \in \Nat.\quad
		e \Step e' \rightarrow
		e' \Eval m \rightarrow e \Eval m
\]
\end{lemma}
\begin{proof}
Assume the two premises $e \Step e'$ and $e' \Eval m$, and proceed by
induction on the structure of the first:
\begin{description}
\item[Rule \eqName{small-plus}:]%{{{%
There are $n_a$ and $n_b$ such that $e \equiv n_a \Add n_b$ and $e' \equiv
n_a + n_b$. As $e'$ is a numeric expression, the only applicable rule for
$e' \Eval m$ is \eqName{big-val}, which implies $m \equiv n_a + n_b$.
Thus the desired conclusion of $e \Eval m$---after substituting for $e$
and $m$---may be satisfied as follows:
\begin{gather*}
\inferrule* [Left=big-plus]
{
	\inferrule* [Left=big-val]
	{ }{n_a \Eval n_a}
	\and
	\inferrule*
	{ }{n_b \Eval n_b}
}{n_a \Add n_b \Eval n_a + n_b}
\end{gather*}
%}}}%
\item[Rule \eqName{small-right}:]%{{{%
There exists $n_a$, $b$ and $b'$ such that $b \Step b'$ with $e \equiv n_a
\Add b$ and $e' \equiv n_a \Add b'$. Substituting for $e'$, the second
assumption becomes $n_a \Add b' \Eval m$, with \eqName{big-plus} as
the only matching rule. This implies the existence of the premises $n_a
\Eval n_a$ and $b' \Eval n_b$,
\begin{gather*}
\inferrule*
{
	\inferrule*
	{ }{n_a \Eval n_a}
	\and
	\inferrule*
	{
		\vdots
	}{b' \Eval n_b}
}{n_a \Add b' \Eval n_a + n_b}
\end{gather*}
for some $n_b$ such that $m \equiv n_a + n_b$. Invoking the induction
hypothesis with $b \Step b'$ and the above derivation of $b' \Eval
n_b$, we obtain a proof of $b \Eval n_b$. The conclusion is satisfied
by the following derivation:
\begin{gather*}
\inferrule*
{
	\inferrule*
	{ }{n_a \Eval n_a}
	\and
	\inferrule* [Right=IH]
	{
		\vdots
	}{b \Eval n_b}
}{n_a \Add b \Eval n_a + n_b}
\end{gather*}
%}}}%
\item[Rule \eqName{small-left}:]%{{{%
This case proceeds in a similar manner to the previous rule, but with $a$,
$a'$ and $b$ such that $a \Step a'$, where $e \equiv a \Add b$ and $e'
\equiv a' \Add b$. Substituting for $e$ and $e'$ in the second assumption
and inspecting its premises, we observe that $a' \Eval n_a$ and $b
\Eval n_b$ for some $n_a$ and $n_b$ where $m \equiv n_a + n_b$:
\begin{gather*}
\inferrule*
{
	\inferrule*
	{
		\vdots
	}{a' \Eval n_a}
	\and
	\inferrule*
	{
		\vdots
	}{b \Eval n_b}
}{a' \Add b \Eval n_a + n_b}
\end{gather*}
Instantiating the induction hypothesis with $a \Step a'$ and $a'
\Eval n_a$ delivers evidence of $a \Eval n_a$. Reusing the second
premise of $b \Eval n_b$ verbatim, we can then derive the
conclusion of $a \Add b \Eval n_a + n_b$:
\begin{gather*}
\inferrule*
{
	\inferrule* [Left=IH]
	{
		\vdots
	}{a \Eval n_a}
	\and
	\inferrule*
	{
		\vdots
	}{b \Eval n_b}
}{a \Add b \Eval n_a + n_b}
\end{gather*}
%}}}%
\end{description}
This completes the proof of $e \Step e' \rightarrow e' \Eval m \rightarrow
e \Eval m$.
\end{proof}

%}}}%

%}}}%

\section{Compiler Correctness}%{{{%

We have given a high-level semantics for the $\Expression$ language in the
previous sections, but in order to run any such program on a real processor,
we need to translate, or \emph{compile} an expression to a sequence of
instructions for some target machine. Given a definition of how the
processor behaves, how do we ensure that the compiler is producing the
correct instructions? When we talk of \emph{compiler correctness}, we mean
that when executed on the target machine, a compiled program ought to
deliver a result coinciding with what one would expect from the high-level
semantics. In this section we define a simple stack-based virtual machine,
along with a compiler for the $\Expression$ language, and demonstrate how we
can prove a corresponding compiler correctness theorem.

\subsection{A Stack Machine and Its Semantics}\label{sec:stack-machine}%{{{%

\def\MS<#1,#2>{\langle #1 ,\; #2 \rangle}

Unlike the previously defined high-level semantics---which specify
transitions between pairs of arbitrarily-sized $\Expression$s---real
processors generally execute a \emph{linear} sequences of instructions, each
mutating the state of the machine in some primitive way. In order to give
such a low-level implementation of our $\Expression$ language, we will make
use of a stack-based virtual machine.

Our stack machine has a stack of natural numbers as its sole form of
storage, and the state of the $\Machine$ at any point may be conveniently
represented as the pair of the currently executing $\Code$, along with the
current $\Stack$,
\begin{align*}
	\Machine & \prod \MS<\Code, \Stack> \\
	\Code & \prod \nil \altn \Instruction :: \Code \\
	\Stack & \prod \nil \altn \Nat :: \Stack
\end{align*}
where $\Code$ comprises a sequence of $\Instruction$s, and $\Stack$
a sequence of values.

Due to the simple nature of the $\Expression$ language, the virtual machine
only requires two $\Instruction$s, both of which operate on the top of the
stack:
\begin{align*}
	\Instruction & \prod \PUSH \Nat \altn \ADD
\end{align*}
The $\PUSH m$ instruction places the number $m$ atop the current stack,
while $\ADD$ replaces the top two values with their sum. Formally, the
semantics of the virtual machine is defined by the $\Exec$
reduction relation, given below:
\begin{gather*}
	\MS<\PUSH m :: c , \sigma>
	\Exec
	\MS<c , m :: \sigma>
		\eqTag{vm-push} \\[2ex]
	\MS<\ADD :: c , n :: m :: \sigma>
	\Exec
	\MS<c , m + n :: \sigma>
		\eqTag{vm-add}
\end{gather*}
As with the previous definition of $\StepS$, we shall write
$\ExecS$ for the transitive, reflexive closure of
$\Exec$:
\begin{gather*}
\inferrule*[right=vm-nil]
{ }{t \ExecS t}
\qquad\qquad
\inferrule*[right=vm-cons]
{a \Exec a' \and a' \ExecS b}
{a \ExecS b}
\end{gather*}

Informally, the difference between the semantics of a virtual machine versus
a high-level semantics is that the reduction rules for the former is simply
a collection of transition rules between pairs of states, and does not make
use of any premises.

%}}}%

\subsection{Compiler}\label{sec:compiler}%{{{%

Given an $\Expression$, a compiler in this context produces some $\Code$
that when executed according to the semantics of the virtual machine just
defined, computes the value of the $\Expression$, leaving the result atop
the current stack. To avoid the need to define concatenation on instruction
sequences and the consequent need to prove various distributive properties,
the definition of $\compile$ below accepts an extra \emph{code continuation}
argument, to which the code for the expression being compiled is prepended.
To compile a top-level expression, we simply pass in the empty sequence
$\nil$. This both simplifies reasoning and results in more efficient
compilers~\cite{hutton07-haskell}. A numeric expression $m$ is compiled to
a $\PUSH m$ instruction, while $a \Add b$ involves compiling the
sub-expressions $a$ and $b$ in turn, followed by an $\ADD$ instruction:
\begin{align*}
	&\compile : \Expression \rightarrow \Code \rightarrow \Code \\
	&\compile\;m\;c = \PUSH m :: c \eqTag{compile-val} \\
	&\compile\;(a \Add b)\;c = \compile\;a\;(\compile\;b\;(\ADD :: c))
		\eqTag{compile-add}
\end{align*}
For example, $\compile\;((1 \Add 2) \Add (4 \Add 8))\;\nil$ produces
the code below,
\begin{gather*}
	\PUSH 1 :: \PUSH 2 :: \ADD :: \PUSH 4 :: \PUSH 8 :: \ADD :: \ADD :: \nil
\end{gather*}
which when executed with an empty initial stack, reduces as follows:
\begin{align*}
	&\MS<\PUSH 1 :: \PUSH 2 :: \ldots, \nil> \\
	&\Exec \MS<\PUSH 2 :: \ADD :: \ldots, 1 :: \nil> \\
	&\Exec \MS<\ADD :: \PUSH 4 :: \ldots, 2 :: 1 :: \nil> \\
	&\Exec \MS<\PUSH 4 :: \PUSH 8 :: \ldots, 3 :: \nil> \\
	&\Exec \MS<\PUSH 8 :: \ADD :: \ldots, 4 :: 3 :: \nil> \\
	&\Exec \MS<\ADD :: \ADD :: \nil, 8 :: 4 :: 3 :: \nil> \\
	&\Exec \MS<\ADD :: \nil, 12 :: 3 :: \nil> \\
	&\Exec \MS<\nil, 15 :: \nil>
\end{align*}

%}}}%

\subsection{Compiler Correctness}%{{{%

By compiler correctness, we mean that given an expression $e$ which
evaluates to $m$ according to a high-level semantics, compiling $e$ and
executing the resultant code on the corresponding virtual machine must
computes the same $m$. Earlier in the chapter, we had shown the mutual
equivalence of the denotational, big-step, and small-step semantics. While
we may freely choose any of these as our high-level semantics, we shall
adopt the big-step semantics, as it makes our later proofs much shorter.

The following equation captures this idea in a rigorous manner:
\begin{gather*}
	e \Eval m
	\quad\leftrightarrow\quad
	\MS<\compile\;e\;\nil, \sigma>
		\ExecS \MS<\nil, m :: \sigma>
\end{gather*}
As $\Eval$ and $\ExecS$ are relations, we necessarily
require the double implication: the $\rightarrow$ direction corresponds to
a notion of \emph{completeness}, and states that the machine must be able to
compute any $m$ that the big-step semantics permits. Conversely, the
$\leftarrow$ direction corresponds to \emph{soundness}, and states that the
machine can only produce values permitted by the big-step semantics.

For the purposes of this proof however, we will need to generalise the
virtual machine on the RHS to an arbitrary code continuation and stack.
\begin{theorem}\label{thm:compiler-correct}
The compiler and virtual machine semantics coincide with the earlier
big-step semantics:
\begin{gather*}
	e \Eval m
	\quad\leftrightarrow\quad
	\forall c, \sigma.\
		\MS<\compile\;e\;c, \sigma>
			\ExecS \MS<c, m :: \sigma>
\end{gather*}
\end{theorem}

\begin{proof}
We shall consider each direction of the double implication separately. In
the forward direction, we assume $e \Eval m$ and proceed on its
structure:
\begin{description}
\item[Rule \eqName{big-val}:]%{{{%
There exists an $n$ such that $e \equiv n$ and $m \equiv n$. Substituting
$n$ for both $e$ and $m$, the conclusion becomes:
\[
		\MS<\compile\;n\;c, \sigma>
			\ExecS \MS<c, n :: \sigma>
	\quad\textrm{, or}\quad
		\MS<\PUSH n :: c, \sigma>
			\ExecS \MS<c, n :: \sigma>
\]
by \eqName{compile-val} in the definition of \compile. The conclusion is
satisfied by applying \eqName{vm-cons} to \eqName{vm-push} and
\eqName{vm-nil}:
\[\inferrule* [Left=vm-cons]
{
	\inferrule* [Left=vm-push]
	{
	}{\MS<\PUSH n :: c, \sigma>
			\Exec \MS<c, n :: \sigma>}
	\and
	\inferrule* [Right=vm-nil]
	{ }{\MS<c, n :: \sigma>
		\ExecS \MS<c, n :: \sigma>}
}{\MS<\PUSH n :: c, \sigma>
			\ExecS \MS<c, n :: \sigma>}
\]
%}}}%
\item[Rule \eqName{big-plus}:]%{{{%
By matching the assumed $e \Eval m$ with the consequent of
\eqName{big-plus}, we see that there exists $a$, $b$, $n_a$ and $n_b$ where
$e \equiv a \Add b$ and $m \equiv n_a + n_b$, such that $a \Eval n_a$ and $b
\Eval n_b$. Substituting for $e$ and $m$, the conclusion becomes
\begin{gather*}
		\MS<\compile\;(a \Add b)\;c, \sigma>
			\ExecS \MS<c, n_a + n_b :: \sigma> \textrm{ , or} \\
		\MS<\compile\;a\;(\compile\;b\;(\ADD :: c)), \sigma>
			\ExecS \MS<c, n_a + n_b :: \sigma>
\end{gather*}
by expanding \compile. Instantiating the induction hypothesis with $a \Eval
n_a$ and $b \Eval n_b$ yields proofs of
\begin{gather*}
	\forall c_a, \sigma_a.\ \MS<\compile\;a\;c_a, \sigma_a> \ExecS \MS<c_a, n_a :: \sigma_a>
		\textrm{ , and} \\
	\forall c_b, \sigma_b.\ \MS<\compile\;b\;c_b, \sigma_b> \ExecS \MS<c_b, n_b :: \sigma_b>
\end{gather*}
respectively. Note that crucially, these two hypotheses are universally
quantified over $c$ and $\sigma$, written with distinct subscripts above to
avoid ambiguity. Now substitute $c_b = \ADD :: c$, $c_a = \compile\;b\;c_b$,
$\sigma_a = \sigma$, $\sigma_b = n_a :: \sigma_a$ and we obtain via the
transitivity of $\ExecS$:
\begin{align*}
	\forall c, \sigma.\
		&\MS<\compile\;a\;(\compile\;b\;(\ADD :: c)), \sigma> \\
	&\ExecS \MS<(\compile\;b\;(\ADD :: c), n_a :: \sigma> \\
	&\ExecS \MS<\ADD :: c, n_b :: n_a :: \sigma>
\end{align*}
A second application of transitivity to \eqName{vm-add} instantiated as
follows,
\begin{gather*}
	\MS<\ADD :: c, n_b :: n_a :: \sigma> \Exec \MS<c, n_a + n_b :: \sigma>
\end{gather*}
gives the required conclusion of:
\begin{align*}
	\forall c, \sigma.\
		&\MS<\compile\;a\;(\compile\;b\;(\ADD :: c)), \sigma> \\
		&\Exec \MS<c, n_a + n_b :: \sigma>
\end{align*}
%}}}%
\end{description}

For the backward direction, we proceed on the structure of $e$:
\begin{description}
\item[Case $e \equiv n$:]%{{{%
Substituting $e$ with $n$, the base case becomes:
\begin{gather*}
	\forall c, \sigma.\
		\MS<\compile\;n\;c, \sigma>
			\ExecS \MS<c, m :: \sigma>
	\rightarrow n \Eval m
	\textrm{ , or} \\
	\forall c, \sigma.\
		\MS<\PUSH n :: c, \sigma>
			\ExecS \MS<c, m :: \sigma>
	\rightarrow n \Eval m
\end{gather*}
Assume the hypothesis and set both $c$ and $\sigma$ to $\nil$ to obtain
$\MS<\PUSH n :: [], []> \ExecS \MS<[], m :: []>$, which must be a single
reduction corresponding to \eqName{vm-push}. Therefore $m$ and $n$ must be
one and the same, and the conclusion of $n \Eval n$ is trivially satisfied
by \eqName{big-val}.
%}}}%
\item[Case $e \equiv a \Add b$:]%{{{%
Substituting $e$ with $a \Add b$ and expanding the definition of \compile,
we need to show that:
\begin{gather*}
	\forall c, \sigma.\
		\MS<\compile\;a\;(\compile\;b\;(\ADD :: c)), \sigma>
			\ExecS \MS<c, m :: \sigma>
	\rightarrow a \Add b \Eval m
\end{gather*}
Now, for both $a$ and $b$, we know that there exists $n_a$ and $n_b$ such
that:
\begin{gather*}
	\forall c_a, \sigma_a.\ \MS<\compile\;a\;c_a, \sigma_a>
		\ExecS \MS<c_a, n_a :: \sigma_a>
	\textrm{ , and} \\
	\forall c_b, \sigma_b.\ \MS<\compile\;b\;c_b, \sigma_b>
		\ExecS \MS<c_b, n_b :: \sigma_b>
\end{gather*}
Substituting for the subscripted $c_a$, $c_b$, $\sigma_a$ and $\sigma_b$ as
we had done in the \eqName{big-plus} case of the first half of this proof,
we obtain:
\begin{gather*}
	\forall c, \sigma.\
		\MS<\compile\;a\;(\compile\;b\;(\ADD :: c)), \sigma>
			\ExecS \MS<c, n_a + n_b :: \sigma>
\end{gather*}
Contrast this with the hypothesis:
\begin{gather*}
	\forall c, \sigma.\
		\MS<\compile\;a\;(\compile\;b\;(\ADD :: c)), \sigma>
			\ExecS \MS<c, m :: \sigma>
\end{gather*}
Since the $\Exec$ reduction relation is deterministic, it must be the case
that $m$ and $n_a + n_b$ are the same. Substituting in $n_a + n_b$ for $m$,
the conclusion becomes $a \Add b \Eval n_a + n_b$---an instance of
\eqName{big-plus}---whose premises of $a \Eval n_a$ and $b \Eval n_b$ are in
turn satisfied by instantiating the induction hypothesis with $\forall c_a,
\sigma_a.\ \MS<\compile\;a\;c_a, \sigma_a> \ExecS \MS<c_a, n_a :: \sigma_a>$
and $\forall c_b, \sigma_b.\ \MS<\compile\;b\;c_b, \sigma_b> \ExecS \MS<c_b,
n_b :: \sigma_b>$.

\TODO{there's more to be said on the determinism of $\Exec$ here---in
contrast with the non-det VMs later---just not sure what it is yet.}

%}}}%
\end{description}
We have shown both the completeness and soundness parts of the compiler
correctness theorem.
\end{proof}

%}}}%

%}}}%

\section{Conclusion}%{{{%

In this chapter, we have shown by example what it means to give the
semantics of simple language in denotational, big-step and small-step
styles. We justify the use of a monoidal model of natural numbers and
addition---with left-to-right evaluation order---as simplification of
monadic sequencing. We then proved the three given semantics to be
equivalent, and demonstrate the use of well-founded induction on the
structure of the reduction rules (that is, rule induction) and of the
syntax. Finally, we have defined a stack-based virtual machine and
a compiler for running programs of the $\Expression$ language, and present
a proof of compiler correctness.

%}}}%

% vim: ft=tex:

